The Spatial and Temporal Domains of Modern Ecology

### 2017-06-13 paper/main.tex
Figure revisions, initial paragraph of discussion section.

Given the growing awareness of scale, expanding data gathering capabilities, and the fact that the most comprehensive (and arguably best-known) analysis of ecological research scales was published in 1989 \cite{tilman_ecological_1989}, it is both timely and important to assess the scales of contemporary ecological investigation. To address this need, we quantified the spatial and temporal domains of empirical observations that were reported within recently (2004-2014) published ecological studies (here domain means the distribution of observations within the spectrum of one or more scale dimensions\footnote{This definition differs slightly from Wiens' \cite{wiens_spatial_1989}, who defined ``domain of scale'' as ``a portion of the scale spectrum within which process-pattern relationships are consistent regardless of scale.''}). Empirical observations are critical for developing and testing the models that explain why ecological patterns vary in time and space \cite{levin_problem_1992, tilman_ecological_1989}, therefore the spatio-temporal domains of observations provide an important indicator of the field's progress towards achieving a holistic, predictive understanding of ecosystems \cite{chave_problem_2013,levin_problem_1992}.

@@

Our analysis was based on a review of 348 papers randomly selected from 42,918 published between 2004-2014 in the top 30 (based on 2012 impact factor) ecology-themed journals. We extracted scale data from 379 observations of ``natural'' (i.e. non-experimentally manipulated) ecological features that were reported within 134 of the reviewed papers (plus an additional 62 that these cited as the source of observations). We excluded experiments because they tend to be of limited extent, duration, and resolution due to their higher logistical costs \cite{tilman_ecological_1989, kareiva_spatial_1988}, and would therefore likely bias our findings towards finer scales.

@@

\noindent \textbf{\emph{Domains}}\\
In terms of resolution, the majority (67\%) of observations were collected in plots of $<$1 m$^2$ resolution, 24\% were collected within plots of 1 m$^2$ up to 1 ha, and the remaining 9\% in plots of $\geq$1 ha (Fig. 1A). The extent of 19\% of observations was $<$10 ha, 23\% covered 10-1,000 ha, 42\% 1000-1,000,000 ha, and 16\% $>$1,000,000 ha (Fig. 1B).

@@

In the temporal dimensions, 37\% of observations were not repeated (Fig. 1C), 17\% were repeated at short intervals (sub-second to daily), 20\% at daily to monthly intervals, 18\% at monthly to yearly intervals, and 8\% at yearly to decadal intervals. Duration was one day or less for 31\% of sampled observations, while 10\% covered one day to one month, 23\% lasted one month to one year, 27\% covered 1-10 years, and 10\% spanned a decade or more (including several paleoecological studies covering centuries to millennia; Fig. 1D).

@@

A similar tendency for duration to increase with interval was also evident amongst temporally replicated observations (Fig. 2D), with the majority of repeated at daily to decadal intervals and spanning $\geq$ 1 month to 1 decade.  The orientation of this concentration shows that interval increases with duration; observations lasting one month to one year tend to have daily to monthly intervals, while those lasting one year to one decade tend to have yearly to decadal intervals. The low densities of observations having sub-daily intervals shows that relatively few high frequency, long duration ecological measurements are undertaken.

To provide further insight into observational domains, we also evaluated the degree to which the scales \emph{represented} by observations differed from those that were actually observed. To make this assessment, we first log$_10$ transformed and then subtracted the values of i) actual extent from extent and ii) actual duration from duration, in order to calculate the magnitude of difference (or decade) between each pair of dimensions for each observation. We then compared how these magnitudes varied in relation to the scale of the actual dimension (Fig. 3). This comparison showed that a majority (81\%) of the assessed observations had actual extents of $\leq$1 ha, which on average were 4 to nearly 8 orders of magnitude smaller than the quantified extent (Fig 3.A).  Actual extent converged with extent above 100,000 ha, but this applies to just 5\% of observations. The actual duration of 64\% of observations was $<$ 1 day, which on average was 3 to nearly 9 orders of magnitude shorter than the time span covered by temporal replicates (Fig. 3B). The two duration measures were approximately equal for the 16\% of observations exceeding one month of actual duration.

@@

\noindent \textbf{\emph{Observational methods}}\\
We classified the method used to collect each observation into several broad categories, which were field methods (manual \emph{in situ} data collection), automated (\emph{in situ}) sensing, remote sensing, other geographic data, and paleo-reconstruction approaches. Field methods were used for 80\% of observations, automated sensing for 12.4\%, remote sensing for 6.3\%, and paleo-reconstruction and other geographic data each for less than 1\%. Using linear regression (weighted by the number of observations per publication year) to assess whether the relative frequency of observing methods changed during the 10 year study period, the use of remote sensing appeared to increase by 1.3\% per year from 2004-2014 (R$^2$ = 0.25, p$<$0.12), and field methods declined by the same percentage (R$^2$ = 0.1, p$<$0.18), although both slopes failed to meet the customary threshold for statistical significance. Automated sensing methods showed no trend over time (SI).

\noindent \textbf{\emph{Potential biases and uncertainties in quantifying scales}}\\
There were several potential methodological aspects that could have influenced our assessment of ecology's spatial and temporal domains. The first stems from our finding that many studies did not precisely report observational scales, which meant that we had to estimate--rather than simply record--these values for most observations (specifically, in 63, 60, and 67\% of cases for resolution, extent, and actual extent, and 36\%, 64\%, and 83\% of cases for interval, duration, and effective duration, respectively). The inevitable estimation errors may have biased our overall findings. However, we attempted to quantify this error by assessing inter-rater disagreement and incorporating this uncertainty into our resampling methodology. The resulting confidence intervals (Fig. 1) suggest that it was unlikely that estimation errors unduly influenced our findings.

@@

\noindent \textbf{Broader implications}\\

Our results show that modern ecological observations most observations are collected at small spatial scales, are either unrepeated or relatively infrequent ($\geq$1 month interval), and in aggregate cover relatively narrow periods of time ($\leq$1 month).

\begin{enumerate}
\item General patterns of scales 
  \subitem{-} General patterns
  \subitem{-} The continued dominance of field-based research, and the limited use of methods that allow larger areas to be comprehensively observed, such as remote sensing.
  \subitem{-} Notable holes: high frequency (sub-daily, daily), high resolution

   \subitem{-} Duration hasn't improved much since Tilman (check this).
   \subitem{-} Evaluate how duration has improved since then. 387/749 of the papers he looked at from the journal Ecology (1977-1984) were mainly experimental or had an experimental component. 99 papers were mainly theoretical or conceptual. Remainder observational. We could compare our results to that here. Of his 623 field based studies (including 180 experiments), $\sim$41\% lasted less than one year,  $\sim$85\% less than three years, most studies $>$10 years and all $>$50 years were paleoecology/chronosequences.
  \subitem{-} Maybe mention here (or probably better later in the discussion), that our decision has the unfortunate consequence of not being directly comparable to previous assessments that included experiments. More of just a caveat or note for the reader who is familiar with the earlier work and their summary statistics
  \subitem But there do seem to be trends increasing extent (and duration, but need to put this above), perhaps due to increasing use of remote sensing 
\item Scales are not well-documented 
  \subitem Furthermore, the unclear documenting of observational scales implies that scale is not a primary concern in much ecological research \cite{chave_problem_2013, wheatley_factors_2009}. (Scale is not a concern that cuts across the entire discipline of ecology).

>>>>

Given the growing awareness of scale, expanding data gathering capabilities, and the fact that the most comprehensive (and arguably best-known) meta-analyses of ecological research scales were published nearly 30 years ago \cite{tilman_ecological_1989,kareiva_spatial_1988}, it is both timely and important to assess the scales of contemporary ecological investigation. To address this need, we quantified the spatial and temporal domains of empirical observations that were reported within recently (2004-2014) published ecological studies (here domain means the distribution of observations within the spectrum of one or more scale dimensions\footnote{This definition differs slightly from Wiens' \cite{wiens_spatial_1989}, who defined ``domain of scale'' as ``a portion of the scale spectrum within which process-pattern relationships are consistent regardless of scale.''}). Empirical observations are critical for developing and testing the models that explain why ecological patterns vary in time and space \cite{levin_problem_1992, tilman_ecological_1989}, therefore the spatio-temporal domains of observations provide an important indicator of the field's progress towards achieving a holistic, predictive understanding of ecosystems \cite{chave_problem_2013,levin_problem_1992}.

@@

Our analysis was based on a review of 348 papers randomly selected from 42,918 published between 2004-2014 in the top 30 (based on 2012 impact factor) ecology-themed journals. We extracted scale data from 378 observations of ``natural'' (i.e. non-experimentally manipulated) ecological features that were reported within 133 of the reviewed papers (plus an additional 62 that these cited as the source of observations). We excluded experiments because they tend to be of limited extent, duration, and resolution due to their higher logistical costs \cite{tilman_ecological_1989, kareiva_spatial_1988}, and would therefore likely bias our findings towards finer scales, while minimizing the impact that new observing methods (e.g. satellite imaging, wireless sensing) may have had in expanding the scales of ecological investigation \cite{turner_remote_2003,pettorelli_satellite_2014,porter_wireless_2005}.

@@

\noindent \textbf{\emph{Domains}}\\
In terms of resolution, the majority (67\%) of observations were collected in plots of $<$1 m$^2$ resolution, 24\% were collected within plots of 1 m$^2$ up to 1 ha, and the remaining 9\% in plots of $\geq$1 ha (Fig. 1A). The extent of 19\% of observations was $<$10 ha, 23\% covered 10-1,000 ha, 11\% 1-10,000 ha, 19\% 10,000-100,000 ha, 27\% 100,000-1,000,000 ha, and 16\% $>$1,000,000 ha (Fig. 1B).

@@

In the temporal dimensions, 37\% of observations were not repeated (Fig. 1C), 17\% were repeated at short intervals (sub-second to daily), 20\% at daily to monthly intervals, 18\% at monthly to yearly intervals, and 8\% at yearly to decadal intervals. Duration was one day or less for 31\% of sampled observations, while 10\% covered one day to one month, 23\% lasted one month to one year, 27\% covered 1-10 years, and 9\% spanned a decade or more (including several paleoecological studies covering centuries to millennia; Fig. 1D).

@@

The majority of temporally replicated observations have daily to decadal intervals and durations of $\geq$ 1 month to 1 decade. The orientation of this concentration shows that interval increases with duration; observations lasting one month to one year tend to have daily to monthly intervals, while those lasting one year to one decade tend to have yearly to decadal intervals. The low densities of observations having sub-daily intervals shows that relatively few high frequency, long duration ecological measurements are undertaken.

To provide further insight into observational domains, we also evaluated the degree to which the scales \emph{represented} by observations differed from those that were actually observed. To make this assessment, we first log$_10$ transformed and then subtracted the values of i) actual extent from extent and ii) actual duration from duration, in order to calculate the magnitude of difference (or decade) between each pair of dimensions for each observation. We then compared how these magnitudes varied in relation to the scale of the actual dimension (Fig. 3). This comparison showed that a majority (81\%) of the assessed observations had actual extents of $\leq$1 ha, which on average were 4 to nearly 8 orders of magnitude smaller than the quantified extent (Fig 3.A).  Actual extent converged with extent $\geq$1 100,000 ha, but this applies to just 5\% of observations. The actual duration of 64\% of observations was $<$ 1 day, which on average was 3 to nearly 9 orders of magnitude shorter than the time span covered by temporal replicates (Fig. 3B). The two duration measures were approximately equal for the 16\% of observations exceeding one month of actual duration.

@@

\noindent \textbf{\emph{Observational methods}}\\
We classified the method used to collect each observation into several broad categories, which were field methods (manual \emph{in situ} data collection), automated (\emph{in situ}) sensing, remote sensing, other geographic data, and paleo-reconstruction approaches. Field methods were used for 80\% of observations, automated sensing for 12.4\%, remote sensing for 6.4\%, and paleo-reconstruction and other geographic data each for less than 1\% each. Using linear regression (weighted by the number of observations per publication year) to assess whether the relative frequency of observing methods changed during the 10 year study period, the use of remote sensing appeared to increase by 1.3\% per year from 2004-2014 (R$^2$ = 0.25, p$<$0.12), and field methods declined by the same percentage (R$^2$ = 0.1, p$<$0.18), although both slopes failed to meet the customary threshold for statistical significance. Automated sensing methods showed no trend over time (SI).

\noindent \textbf{\emph{Potential biases and uncertainties in quantifying scales}}\\
There were several potential methodological aspects that could have influenced our assessment of ecology's spatial and temporal domains. The first stems from our finding that many studies did not precisely report observational scales, which meant that we had to estimate--rather than simply record--these values for most observations (specifically, in 63, 60, and 69\% of cases for resolution, extent, and actual extent, and 36\%, 64\%, and 83\% of cases for interval, duration, and effective duration, respectively). The inevitable estimation errors may have biased our overall findings. However, we attempted to quantify this error by assessing inter-rater disagreement and incorporating this uncertainty into our resampling methodology. The resulting confidence intervals (Fig. 1) suggest that it was unlikely that estimation errors unduly influenced our findings.

@@

\noindent \textbf{Broader implications}\\
Our results suggest that the spatial and temporal domains of modern ecological observations are fairly narrow, and are collected primarily with conventional field techniques. Spatially, the majority of observations have grains of $<1$ m$^2$ and extents of $<$10,000 ha (Fig 1A;B). In the temporal domains, most observations are either un-replicated snapshots, or of low frequency ($\geq$1 month interval; Fig. 1C) and relatively short duration ($leq$1 year; Fig. 1D; 2D). Comparisons between observational dimensions revealed that larger extents and are associated with larger plot sizes (Fig. 2C), while longer durations are associated with longer intervals (Fig. 2D). The latter association presumably reflects a cost-imposed tradeoff between sampling frequency and temporal duration that is characteristic of field-based data collection methods. This same tradeoff is also responsible for the inverse relationship between resolution and interval that is the most obvious feature within that domain space (Fig. 1A).

Comparing these general findings to those of earlier related studies, it appears that modern ecology's observational scales have not changed greatly in the past few decades.

These general results suggest that the scale domains of modern ecology have not changed greatly since the 1980s.  Porter's study.

Comparing these primary tendencies to earlier efforts, it seems that not much has changed in ecological

Taken collectively, these

\begin{enumerate}
\item General patterns of scales 
  \subitem{-} General patterns
  \subitem{-} The continued dominance of field-based research, and the limited use of methods that allow larger areas to be comprehensively observed, such as remote sensing.
  \subitem{-} Notable holes: high frequency (sub-daily, daily), high resolution 
   \subitem{-} Change since Tilman/Kareiva
   \subitem{-} Duration hasn't improved much since Tilman (check this).
   \subitem{-} Evaluate how duration has improved since then. 387/749 of the papers he looked at from the journal Ecology (1977-1984) were mainly experimental or had an experimental component. 99 papers were mainly theoretical or conceptual. Remainder observational. We could compare our results to that here. Of his 623 field based studies (including 180 experiments), $\sim$41\% lasted less than one year,  $\sim$85\% less than three years, most studies $>$10 years and all $>$50 years were paleoecology/chronosequences.
  \subitem{-} Maybe mention here (or probably better later in the discussion), that our decision has the unfortunate consequence of not being directly comparable to previous assessments that included experiments. More of just a caveat or note for the reader who is familiar with the earlier work and their summary statistics
  \subitem But there do seem to be trends increasing extent (and duration, but need to put this above), perhaps due to increasing use of remote sensing 
\item Scales are not well-documented 
  \subitem Furthermore, the unclear documenting of observational scales implies that scale is not a primary concern in much ecological research \cite{chave_problem_2013, wheatley_factors_2009}. (Scale is not a concern that cuts across the entire discipline of ecology).
